# Background
After Hurricane Hugo made impact on Puerto Rico in 1989, Schaefer et al. ([2000](https://doi.org/10.1017/s0266467400001358)) studied the long-term stream water quality in Bisley, Puerto Rico.

This project aims to replicate Figure 3 from Schaefer et al. ([2000](https://doi.org/10.1017/s0266467400001358)). We focused on potassium, nitrate, magnesium, calcium, and ammonium at four sites. The four sites are Puente Roto Mameyes (PRM), Quebrada one-Bisley (BQ1), Quebrada two-Bisley (BQ2), and Quebrada three-Bisley (BQ3).

# Data
Data is accessed through the [Environmental Data Initiative](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-luq.20.4923064), by International Institute of Tropical Forestry(IITF), USDA Forest Service.
```{r, output = FALSE}
library(tidyverse)
bq1 <- read_csv("../data/QuebradaCuenca1-Bisley.csv")
bq2 <- read_csv("../data/QuebradaCuenca2-Bisley.csv")
bq3 <- read_csv("../data/QuebradaCuenca3-Bisley.csv")
prm <- read_csv("../data/RioMameyesPuenteRoto.csv")
```

# Methods
The watersheds in this study were all located in the Rio Mameyes drainage basin within Luquillo Experimental Forest (LEF) in northeastern Puerto Rico.  
From the EDI database, we downloaded the dataset for Puente Royo Mameyes and Bisley Quebrada 1, 2, and 3.
When reading in the data, we defined a function that calculated the ion concentration level for a 9-week moving average by site for potassium, calcium, magnesium, ammonium, and nitrate. 

```{r, output = FALSE}
source("../R/moving-average.R")

bq1_mvavg <- moving_average("BQ1", bq1)
bq2_mvavg <- moving_average("BQ2", bq2)
bq3_mvavg <- moving_average("BQ3", bq3)
prm_mvavg <- moving_average("PRM", prm)

big_frame <- bind_rows(bq1_mvavg, bq2_mvavg, bq3_mvavg, prm_mvavg)
glimpse(big_frame)

big_frame_longer <- big_frame |> 
  pivot_longer(
    cols = "k_mgl":"no3_ugl",
    names_to = "ions",
    values_to = "concentration"
  )
```

# Results
```{r}
ggplot(
  data = big_frame_longer,
  mapping = aes(
    x = window_start,
    y = concentration,
    color = site
  )
) +
  geom_line() +
  labs(
    title = "Moving average of ions in stream water",
    x = "Years"
  ) +
  facet_wrap(
    ~ions,
    scales = "free",
    ncol = 1,
    strip.position = "left",
    labeller = as_labeller(c(k_mgl="K mg l^-1",
                            ca_mgl="Ca mg l^-1",
                            mg_mgl="Mg mg l^-1",
                            nh4_ugl="NH4-N ug l^-1",
                            no3_ugl="NO3-N ug l^-1"))
  ) +
  theme_bw()
```

## References
- McDowell, William H., and USDA Forest Service. International Institute Of Tropical Forestry (IITF). 2024. “Chemistry of Stream Water from the Luquillo Mountains.” Environmental Data Initiative. [doi.org/10.6073/PASTA/F31349BEBDC304F758718F4798D25458.](https://doi.org/10.6073/PASTA/F31349BEBDC304F758718F4798D25458)
- Schaefer, Douglas. A., William H. McDowell, Fredrick N Scatena, and Clyde E. Asbury. 2000. “Effects of Hurricane Disturbance on Stream Water Concentrations and Fluxes in Eight Tropical Forest Watersheds of the Luquillo Experimental Forest, Puerto Rico.” Journal of Tropical Ecology 16 (2): 189–207. [doi.org/10.1017/s0266467400001358.](https://doi.org/10.1017/s0266467400001358)