# Hurricane Hugo (1989) impact on Puerto Rico's Water Quality

## Background
After Hurricane Hugo made impact on Puerto Rico in 1989, Schaefer et al. ([2000](https://doi.org/10.1017/s0266467400001358)) studied the long-term stream water quality in **Bisley, Puerto Rico**.

This project aims to **replicate Figure 3 from Schaefer et al. ([2000](https://doi.org/10.1017/s0266467400001358))**. We focused on potassium (K), nitrate (NO3-N), magnesium (Mg), calcium (Ca), and ammonium (NH4-N) at four sites. The four sites are Puente Roto Mameyes (PRM), Quebrada one-Bisley (BQ1), Quebrada two-Bisley (BQ2), and Quebrada three-Bisley (BQ3).

![Concentration of (a) potassium, (b) nitrate, (c) magnesium, (d) calcium, and (e) ammonium sampled in Bisley, Puerto Rico after Hurricane Hugo according to Schaefer et al. ([2000](https://doi.org/10.1017/s0266467400001358)).](schaefer_fig3.png){fig-align="center"}

## Data
Data is accessed through the [Environmental Data Initiative](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-luq.20.4923064), by International Institute of Tropical Forestry (IITF), USDA Forest Service.

We extracted the Site ID, Sample Date, and the concentration of potassium, nitrate, magnesium, calcium, and ammonium from Puente Royo Mameyes and Bisley Quebrada 1, 2, and 3.
```{r, output = FALSE, echo = FALSE}
library(tidyverse)
df <- read_csv("../output/allsite_mvavg.csv")
df$ions <- factor(df$ions, levels=c("k_mgl","no3_ugl","mg_mgl","ca_mgl","nh4_ugl"))
df$site <- factor(df$site, levels=c("PRM","BQ1","BQ2","BQ3"))
```

## Methods
The watersheds in this study were all located in the Rio Mameyes drainage basin within Luquillo Experimental Forest (LEF) in northeastern Puerto Rico.  

We calculated the ion concentration level for a 9-week moving average by site for potassium, calcium, magnesium, ammonium, and nitrate from Jan 1, 1988 to Jan 1, 1995.  

- A _moving average_ is taking all the values within a time value (9 weeks in this case) and calculating the average value within that period.

## Results
In R, we utilize `ggplot()` in the R library `tidyverse` to replicate Figure 3 from Schaefer et al. ([2000](https://doi.org/10.1017/s0266467400001358)).  
The code is shown below:
```{r}
#| warning: false
#| echo: false
#| fig-width: 7.5
#| fig-height: 8
ggplot(
  data = df,
  mapping = aes(
    x = window_start,
    y = concentration,
    linetype = site
  )
) +
  geom_line() +
  labs(
    title = "Moving average of ions in stream water",
    x = "Years"
  ) +
  facet_wrap(
    ~ions,
    scales = "free",
    ncol = 1,
    strip.position = "left",
    labeller = as_labeller(c(k_mgl="K mg/l",
                            ca_mgl="Ca mg/l",
                            mg_mgl="Mg mg/l",
                            nh4_ugl="NH4-N ug/l",
                            no3_ugl="NO3-N ug/l"))
  ) +
  theme_bw() +
  geom_vline(xintercept = ymd("1989-09-18"), linetype = "dashed") +
  scale_linetype_manual(values = c("PRM" = "solid", "BQ1" = "dotted", "BQ2" = "dashed", "BQ3" = "dotdash"))
```
_Concentration of (a) potassium, (b) nitrate, (c) magnesium, (d) calcium, and (e) ammonium sampled in Bisley, Puerto Rico after Hurricane Hugo based on data from the Environmental Data Initiative._

The data availability from EDI is different from Schaefer et al. ([2000](https://doi.org/10.1017/s0266467400001358)). While EDI has more data available for the magnesium readings, the ammonium data collection stopped after mid 1988.  
Aside from the differences in data availability, we were able to successfully replicate Figure 3 from Schaefer et al. ([2000](https://doi.org/10.1017/s0266467400001358)). However, the lines from my recreated graph is not as smooth.

### Source
- **Author**: Kelly Loo, [kloo@ucsb.edu](kloo@ucsb.edu)
- **Collaborators**: Oliver Witt, Oliva Knapp
- **Date Created**: Aug 24, 2027

### References
- McDowell, William H., and USDA Forest Service. International Institute Of Tropical Forestry (IITF). 2024. “Chemistry of Stream Water from the Luquillo Mountains.” Environmental Data Initiative. [doi.org/10.6073/PASTA/F31349BEBDC304F758718F4798D25458.](https://doi.org/10.6073/PASTA/F31349BEBDC304F758718F4798D25458)
- Schaefer, Douglas. A., William H. McDowell, Fredrick N Scatena, and Clyde E. Asbury. 2000. “Effects of Hurricane Disturbance on Stream Water Concentrations and Fluxes in Eight Tropical Forest Watersheds of the Luquillo Experimental Forest, Puerto Rico.” Journal of Tropical Ecology 16 (2): 189–207. [doi.org/10.1017/s0266467400001358.](https://doi.org/10.1017/s0266467400001358)